Case-based Arabic Morphological Analyzer for Arabic Text Mining

### 2010-12-19 trunk/tex/rp.tex
comments and add rb

The almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}
covers most academic research work on Arabic text mining.
It summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While machine translation is of gneral interest, 
an Arabic user is more interested in 
information hard to discover manually.
Examples are {the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness,
the relation
between stock prices and political news, or the 
relation between interrogation reports with suspects
and investigative reports from the crime scenes.

We propose to build a framework for
%{\em relational text mining 
%of Arabic using morphological and case-based analysis}. 
{\em case-based Arabic text analysis using relations and morphology}(CATARM). 
CATARM takes as input several sets of documents, each of a specific
type, and a user query. % expressed in a visual language.
The query, expressed in a visual language, allows the user 
to build a flow of Arabic text mining
components and discover desired relations between documents. 
CATARM compiles the query into a netlist of computational 
components. It also uses features extracted from the query and 
the document sets to simplify the computational 
components. 
%The main CATRAM computational components will be finite 
%state machines (FSM) that express a linguistic model of the Arabic 
%language and that can be parametrized and relaxed to fit a case 
%study.

\novocalize
%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

@@

We find evidence to our hypothesis in~\cite{Maamouri:10} where 
a new corpus was added to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}. 
The addition required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new genre of data challenged the existing data processing and 
annotation guidelines. 
Alternative guidelines came to place where the high level NLP 
task interacted with the morphological analyzer. 
This led to refining SAMA~\cite{Kulick:10}. 
The work of Habash~\shortcite{Habash:06} reports that different 
types of morphological analyses behave differently on the same case 
study. 
For one, the analyzers may find different solutions. 
The analyzers may present the solutions in different orders. 
The presented solutions and the order of presentation may not be 
appropriate for the case study. 
We find strong evidence in our preliminary results 
that a case-based morphological analyzer that allows a case-based 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

@@

{\bf Hypothesis 2.} This approach does not tend directly to 
the Arabic users who write their 
queries in Arabic and think in Arabic.
We hypothesize that Arabic queries use rules and 
relations similar to the Arabic text under investigation, 
and thus techniques native to the Arabic language will do better 
in Arabic text analysis.

We find evidence in the work of the CADIM \noTrRL{q-adim} 
project~\cite{Col09} at the University of Columbia to support our 
hypothesis. They observe that Arabic texts used as training 
data for common NLP techniques proved ineffective. 
They built semi-automatic models to address the problem of automatic 
speech recognition in Arabic.

@@

ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
approach for morphological analysis. 
Realizational morphology considers the form of the word 
\noTrRL{.sarf}  instead of its segments. 
Formal rules and formulas express the relations between 
the root of the lexeme and the fully inflected word. 
While this is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04},
and it benefits from features unique to the Arabic language
captured by the formal rules, 
it is computationally expensive and enumerates all the possible
solutions.

{\bf Goals.} CATARM 
tends to relational Arabic text mining using morphological analysis.
{\bf (1.)} CATARM will build a novel parametrized 
linguistic computational model (LCM) of Arabic. 
The LCM will be implemented using efficient structures. 
 {\bf (2.)} The LCM will be amenable to Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the relational user queries.
{\bf (3.)} The LCM will be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as 
hidden Markov models (HMM)  and 
latent semantic indexing~\cite{LSI89}.
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from text 
documents and uses set arithmetic and graph analysis techniques
to establish relations between the graphs. {\bf (5.)} CATARM will 
solve at least three interesting case studies such as those
presented in Sections~\ref{s:design:lit} and~\ref{s:design:sec}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will build a system using CATARM to check the strength
of the relations between Internet documents viewed by a minor and
a set of Arabic text documents containing unrated or unwanted 
material. If the relation is strong, the system will flag access to
the document. 
All components and parts built during the development of solutions
to the case studies will be generalized and added to CATARM as 
library computational components. 
\end{itemize}

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits in cases similar to the above mentioned 
case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.

The use of Arabic text mining  will 
have a high impact on the prosperity and safety of the Arab world 
in particular and the world in general.
%For example, having the tools to check the authenticity of a hadith 
%at a click of a button may save youth from the hands of extreme 
%religious leaders.
For example, extracting a relation between two criminals may 
uncover a third criminal who is ready to commit a crime.
Finding inconsistencies in claims may curb down fraud and corruption 
in the insurance sector.

@@

@@

A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
We consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into
its content or \RL{matn} and its \noArRL{sanad} or the
chain of narrators.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names 
and once
connected together they form complex names of narrators. 
Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
name is \noTrRL{s`yd} as the word in between 
\noTrRL{bn} (son of) indicates.

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.7\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
\transtrue

@@

\begin{figure}[tb]
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/narrator_chain_output.pdf}}
\caption{Directed acyclic graph representing a partial order 
    relation between narrators extracted using ATSarf}
\label{f:narrators} 
}
\end{figure}

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) between
narrators.
We automatically extracted the POR from hadith text 
documents.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localizing narrators in biography documents and
segmenting biography documents.

@@

\subsection{PI related work}

The Co-LPI, Rehab Duwairi, has several years of experience 
working on data mining for Arabic documents. 
She initially targeted Arabic text categorization where a 
data set of 1000 documents where used to train and test 
classifiers~\cite{Duw06,Duw07}.
Word morphology comes to the surface when extracting the 
feature vectors of documents. 
In that work, words were stemmed before using them as features. 
Also, word frequency and inverted term frequency were used to 
quantify the importance of features.

@@

The Lead PI worked in the area of software arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of literature heritage~\cite{Zar06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

@@

\subsection{Literature authentication case study}
\label{s:design:lit}

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.75\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
} 
\caption{CATARM diagram for the literature case study }
\label{f:literature}
}
\end{figure}

The diagram in Figure~\ref{f:literature} describes our 
high level vision for what the literature case study
will look like in the CATARM framework. 
The boxes represent 
computational blocks provided by CATARM, 
ovals represent data blocks which are mainly the 
inputs and the outputs of the computational blocks. 
Dotted data blocks represent intermediate results,
and solid and shaded blocks represent the initial input 
and the final output respectively.

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/security_case-study.pdf}
}
\caption{CATARM diagram for the security case study}
\label{f:security}
}
\end{figure}

@@

An MA with date, place, and name extraction controller extracts
names, dates and places from the investigation reports. 
We pass the dates to a sorting component to produce $G_D$ 
a partial order graph between the dates. 
We look up the names in the personal records database and find out
$G_N$ a family dependency graph between the names. 
We pass the places to a mapper that builds another $G_G$,
a graph relating the places using a geographical database.

@@

@@

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morphological solution for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

@@

\begin{figure}
\center{
\resizebox{.4\columnwidth}{!}
{\input{figs/hadith.pdftex_t}
}
\caption{State-machine used to detect the chain of narrators}
\label{f:statemachine}
}
\end{figure}

{\bf Case-based morphological analyzer.}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

@@

{\bf Linguistic computational model of Arabic.}
We think that a linguistic computational model for the Arabic
language will perform better in applications where
relations between Arabic text documents are involved. 
Such a model helps capture features unique and specific 
to the Arabic language and make use of that in the analysis.

ElixirFM~\cite{Otakar:07} builds on the lexicon
of the Buckwalter analyzer and the annotations of the 
Prague Arabic dependency databank to build a formal 
computational model of the Arabic language.
The models is a set of formulas and rules. 
ElixirFM uses the computational model in a functional
morphology framework that uses Haskell and functional
programming. 
The analyzer works by generating morphological trees of 
decisions and then uses a disambiguation decision procedure 
to annotate the trees with contextual elements.

@@

{\bf User query language. }
In Figure~\ref{f:statemachine} we showed an FSM 
that controls the morphological analysis and generates a
list of chains of narrators out of hadith text.
We think that it is overwhelming to ask a user who is not 
versed in FSM and in computational sciences in general 
to design such a controller.

@@

\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
\label{f:chain}
}
\end{figure}

@@

{\bf Simplify LCM based on the query.}
Given
L$L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to over-approximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

This task is not trivial though, since $\rho_q$ may contain
states and transitions that are not always morphemes and 
transition between morphemes. 
We will consider a hierarchical methodology that follows 
the hierarchical structure of the Arabic language expressed
in $L_A$. For example,
we may consider sets of connected states in $\rho_q$ 
containing a single morpheme as a state in one hierarchy level 
in $L_A$.
We will perform this simplification before starting the 
morphological analysis since we already have to analyze
the query $q$ and compute $\rho_q$ before we start
the analysis.

{\bf Simplify LCM and MA based on set of documents.}
We will analyze the set of input documents given to a
morphological analyzer component and use the analysis
to further improve the morphological analysis. 
We will extract statistical features from these documents 
to help order the solutions considered by the analyzer
to consider the most probable first.

We will consider the use of HMM models 
built on the fly while processing the documents.
We will also consider the use of
latent semantic indexing~\cite{LSI89} techniques. 
We are considering these two techniques as they are 
both language independent and can both be computed and
adapted on the fly.

{\bf Entity and relational extractor.}
We will build a set of entity and relational extractors. 
The entities can be proper names, dates, locations,
and qualification adjectives. 
We will explore the use of pairing techniques to establish 
relations between the extracted entities.
The relations will form graphs as in the case studies presented
so far. The nodes represent the entities and the previously 
extracted relations and the edges represent the newly discovered 
relations.

@@

{\bf Statistical components.}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.

>>>>

The almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}
covers most academic research work on Arabic text mining.
It summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While machine translation is of general interest, 
an Arabic user is more interested in 
information hard to discover manually.
Examples are the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness,
the relation
between stock prices and political news, or the 
relation between interrogation reports with suspects
and investigative reports from the crime scenes.

We propose to build a framework for
%{\em relational text mining 
%of Arabic using morphological and case-based analysis}. 
{\em case-based Arabic text analysis using relations and morphology} (CATARM). 
CATARM will apply novel algorithms and tools
to answer queries seeking relations amongst
Arabic text documents.
CATARM takes as input several sets of documents, each of a specific
type, and a user query. % expressed in a visual language.
The query, expressed in a visual language, allows the user 
to build a flow of Arabic text mining
components and discover desired relations between documents. 
CATARM compiles the query into a netlist of computational 
components. It also uses features extracted from the query and 
the document sets to simplify the computational 
components. 
%The main CATRAM computational components will be finite 
%state machines (FSM) that express a linguistic model of the Arabic 
%language and that can be parametrized and relaxed to fit a case 
%study.

\novocalize
%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e. beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

@@

We find evidence to our hypothesis in~\cite{Maamouri:10} where 
a new corpus was added to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}. 
The addition required compliance with the DARPA GALE program 
including several NLP tasks such as English translation 
annotations and word-level alignment of Arabic and English. 
The new genre of data challenged the existing data processing and 
annotation guidelines. 
Alternative guidelines came to place where the high level NLP 
task interacted with the morphological analyzer. 
This led to refining SAMA~\cite{Kulick:10}. 
The work of Habash~\cite{Habash:06} reports that different 
types of morphological analyses behave differently on the same case 
study. 
For one, the analyzers may find different solutions. 
The analyzers may present the solutions in different orders. 
The presented solutions and the order of presentation may not be 
appropriate for the case study. 
We find strong evidence in our preliminary results 
that a case-based morphological analyzer that allows a case-based 
controller to intervene at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

@@

{\bf Hypothesis 2.} 
The preprocessing approach does not tend directly to 
the Arabic users who write their 
queries in Arabic and think in Arabic.
We hypothesize that Arabic queries use rules and 
look for relations similar in nature to the structure of the Arabic language 
and the text under investigation.
Thus techniques native to the Arabic language will do better 
in Arabic text analysis.

The work of the CADIM \noTrRL{q-adim} 
project~\cite{Col09} at the University of Columbia supports our 
hypothesis. They observe that Arabic texts used as training 
data for common NLP techniques proved ineffective. 
They built semi-automatic models to address the problem of
Arabic speech recognition.

@@

ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
approach for morphological analysis
and considers the form of the word 
\noTrRL{.sarf}  instead of its segments. 
Formal rules express the relations between 
the root of the lexeme and the fully inflected word. 
While the rules capture features unique to the 
the Arabic language and its linguistic description~\cite{Badawi:04},
the approach is computationally expensive and enumerates 
all the possible solutions.

{\bf Goals.} 
CATARM tends to relational Arabic text mining using 
{\bf (1.)} a novel case-based morphological analyzer.
{\bf (2.)} CATARM will build a novel and efficient 
linguistic computational model (LCM) for Arabic. 
%The LCM will be implemented using efficient structures. 
The LCM will be amenable to novel 
{\bf (3.)} Arabic specific simplifications 
inferred from the correspondence between the formal Arabic rules
and the user queries, and 
{\bf (4.)} language independent simplifications
inferred from statistical features %extracted from the documents 
using language independent techniques such as 
hidden Markov models (HMM)~\cite{JanHMM06}
and latent semantic indexing (LSI)~\cite{LSI89}.
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (5.)} CATARM extracts semantic graphs 
and uses set arithmetic and graph analysis techniques
to establish relations between the graphs. 
{\bf (6.)} CATARM will solve the case studies 
presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.

@@

\begin{itemize}\itemsep0pt
\item CATARM will build a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item CATARM will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item CATARM will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\item We will use CATARM to solve several realistic case studies.
We will look at the hadith literature and build a system using
CATARM to relate the hadith literature to the biographies of the
narrators and extract an authenticity graph for the narrators.
We will look at investigation and interrogation security reports
and build a CATARM system that looks for relations between suspects
in the reports. 
We will look at clinical reports and relate them to pharmaceutical 
records and find relations between sicknesses and prescribed medication. 
Finally, we will use CATARM to provide content-based parental control
for Arabic text documents. We will look for 
relations between documents requested by a minor and
a set of undesired Arabic documents. 
If the relation is strong, we will flag access to
the documents.

We further explain our approach with examples and discuss 
how we will obtain the documents 
in Sections~\ref{s:design:lit}--\ref{s:design:med}.
We will generalize all components built during the development 
of the case studies and we will add them to 
the libraries of CATARM. 
%library computational components. 
\end{itemize}

@@

CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits.% in cases similar to the above mentioned 
%case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.
CATARM may be used to flag spammers 
posting biased reviews in Arabic web forums.

The use of Arabic text mining  will 
have a high impact on the prosperity and safety of the Arab world 
in particular and the world in general.
%For example, having the tools to check the authenticity of a hadith 
%at a click of a button may save youth from the hands of extreme 
%religious leaders.
For example, extracting a relation between two criminals may 
uncover a third criminal ready to commit a crime.
Finding inconsistencies in claims may curb down fraud 
in the insurance sector.

@@

\transfalse
%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.8\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
\transtrue

@@

A \RL{.hady_t} is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
Figure~\ref{f:exhadith} shows an example~\noArRL{.hady_t} in Arabic with its 
transliteration and translation. 
The authentication of a \noArRL{.hady_t} highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
We consider the problem of automatically segmenting
a \noArRL{.hady_t} book into narrations, then segmenting each 
narration into
its content or \RL{matn} and its \noArRL{sanad} or the
chain of narrators.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books. 
For example, the boxes in Figure~\ref{f:exhadith} are proper names 
and once
connected together they form complex names of narrators. 
Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
name is \noTrRL{s`yd} as the word in between 
\noTrRL{bn} (son of) indicates.

@@

\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/narrator_chain_output.pdf}}
\caption{Directed acyclic graph representing a partial order 
    relation between narrators extracted using ATSarf}
\label{f:narrators} 
}
\end{figure}

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from three books of hadith 
selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}
\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
This partial order graph is instrumental to automate
localizing narrators in biography documents and
segmenting biography documents.

@@

\subsection{Related work of principal investigators}

The Co-LPI, Rehab Duwairi, has several years of experience 
working on data mining for Arabic documents. 
She initially targeted Arabic text categorization where a 
data set of 1,000 documents was used to train and test 
classifiers~\cite{Duw06,Duw07}.
Word morphology comes to the surface when extracting the 
feature vectors of documents. 
In that work, words were stemmed before using them as features. 
Also, word frequency and inverted term frequency were used to 
quantify the importance of features.

@@

The Lead PI worked in the area of software Arabization since 1996 
as he worked on developing Arabization modules for Windows NT 
and an Arabic string manipulation library~\cite{Zar96}. 
The Lead PI is also involved in working on an open source 
project that aims at comprehensive automation of authenticity 
checks against a corpus of hadith literature~\cite{Zar06}.
The Lead PI is leading an effort at AUB to introduce 
Computational Arabic in the curriculum of the ECE program. 
He is also leading a team of students in an effort to place 
the infrastructure needed for this project to start.

@@

We describe CATARM using four case studies and then move
to describe the different tools needed to build
CATARM.

\subsection{Hadith authentication case study}
\label{s:design:lit}

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/literature_case-study1.pdf}
} 
\caption{CATARM diagram for the hadith case study }
\label{f:literature}
}
\end{figure}

The diagram in Figure~\ref{f:literature} describes our 
high level vision for what the hadith case study
will look like in the CATARM framework. 
The boxes represent 
computational blocks provided by CATARM, 
ovals represent data blocks which are mainly the 
inputs and the outputs of the computational blocks. 
Dotted data blocks represent intermediate results,
and solid and shaded blocks represent the initial input 
and the final output respectively.

@@

%\begin{figure}[tb]
\begin{figure}
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/security_case-study.pdf}
}
\caption{CATARM diagram for the security case study}
\label{f:security}
}
\end{figure}

@@

A morphological analyzer with date, place, and name extraction controller extracts
names, dates and places from the investigation reports. 
We pass the dates to a sorting component to produce $G_D$ 
a partial order graph between the dates. 
We look up the names in the personal records database and find out
$G_N$ a family dependency graph between the names. 
We pass the places to a mapper that builds another $G_G$,
a graph relating the places using a geographical database.

@@

As such reports contain sensitive information and our contacts
in the Lebanese security declined direct access, 
we will start the security case study by studying dummy documents that 
represent both interrogative and investigative types of documents. 
We collected a small amount (in the order of tens)
of similar documents published in the news in the last two years
from cases that were brought to public court or from high profile
cases that leaked to the media.
We will work on augmenting this collection further to have a large
enough set of reports (in the order of hundreds). 
We will mutate the entities in these documents and seed the mutated
set with relations we wish to discover and that are hard to discover
manually. 
Later on, we will use our contacts with the Lebanese and Qatari 
security authorities
to validate our system on their real data.

\subsection{Network content case study}
\label{s:design:net}

Parental controls help parents allow their kids to benefit from the internet 
without fearing the danger of violent and obscene material. 
These controls partially rely on processing the content to detect and rate 
the target material. 
Control mechanisms for Arabic traffic as currently mostly manual. 
In this case study we will detect the bad content. 
Given a set of undesired internet content 
and a set of requests by internet users, 
CATARM can discover whether there is a relation between 
the requested documents and the undesired content and flag the 
requested document with strong enough relation.

\subsection{Health case study}
\label{s:design:med}

Clinical reports contain diagnosis information, observations about 
patient progress, prescriptions, and patient response to prescribed 
medications. 
In countries, like Lebanon, where medical doctors graduate and specialize 
in several distinct practices around the world it is interesting to 
check for common practices and emerging trends to gain insight
about local and regional health and diagnosis features. 
It is also interesting to relate 
pharmaceutical 
records to clinical reports to check on 
market consistency and on self prescribed medicine 
practices.
Such a study can uncover serious tendencies in uncontrolled
pharmaceutical markets.
It can uncover what kind of sicknesses and what kind of medicines
are used sensitive medicines are used without 
appropriate medical supervision.

We will work with our contacts at the American University
Hospital (AUH) at Beirut to analyze this data and relate it
to nearby pharmaceutical records.

@@

\subsection{Case-based morphological analyzer}
\label{s:design:ma}

Morphological analyzers for the Arabic language
such as Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} take an Arabic word as input
and produce several valid morphological solutions for the word. 
Each solution is a sequence of morphemes that form the word.
Each morpheme is associated with several tags such as 
morphological, part of speech, grammar, and semantic tags. 
These analyzers are essential in several text analysis tools such 
as spell checkers as well as NLP frameworks~\cite{Col09}.
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

@@

\begin{figure}
\center{
\resizebox{.5\columnwidth}{!}
{\input{figs/hadith.pdftex_t}
}
\caption{State-machine used to detect the chain of narrators}
\label{f:statemachine}
}
\end{figure}

{\bf Case-based morphological analyzer.~~}
We experimented with several morphological analyzers.
We think that using finite state machines for morphological analysis
can be implemented in a more efficient way compared to other 
approaches. 
For example the concatenative rules used in Buckwalter~\cite{Tim04}
can be encoded in three FSM with linear execution time using 
the trie~\cite{Aoe:89} data structure.

@@

{\bf Linguistic computational model of Arabic.~~}
We think that a linguistic computational model for the Arabic
language will perform better in applications where
relations between Arabic text documents are involved. 
Such a model helps capture features unique and specific 
to the Arabic language and make use of that in the analysis.

ElixirFM~\cite{Otakar:07} builds on the lexicon
of the Buckwalter analyzer and the annotations of the 
Prague Arabic dependency databank to build a formal 
computational model of the Arabic language.
The model is a set of formulas and rules. 
ElixirFM uses the computational model in a functional
morphology framework that uses Haskell and functional
programming. 
The analyzer works by generating morphological trees of 
decisions and then uses a disambiguation decision procedure 
to annotate the trees with contextual elements.

@@

{\bf User query language.~~}
In Figure~\ref{f:statemachine} we showed an FSM 
that controls the morphological analysis and generates a
list of chains of narrators out of hadith text.
We think that it is overwhelming to ask a user who is not 
versed in FSM and in computational sciences in general 
to design such a controller.

@@

\begin{figure}
\center{
\resizebox{.9\columnwidth}{!}
{\includegraphics{figs/detailed_submachine.pdf}
}
\caption{CATARM detailed diagram for the chain of narrators extractor}
\label{f:chain}
}
\end{figure}

@@

{\bf Simplify LCM based on the query.~~}
Given $L_q$ the query language, 
$L_A$ the Arabic linguistic
computational model, and  $q$ a user query
we compute $\rho_q$, the grammar path that matches 
the query $q$ in $L_q$. 
We plan to inspect methods to over-approximate 
the morphemes and the transitions in $L_A$ 
that can be satisfied by morphemes and transitions
in $\rho_q$. 
This can simplify the number of states we need to 
look at under $L_A$.

This task is not trivial though, since $\rho_q$ may contain
states and transitions that are not always morphemes and 
transitions between morphemes. 
We will consider a hierarchical methodology that follows 
the hierarchical structure of the Arabic language expressed
in $L_A$. 
Consider $\Phi$ the state machine representing the grammar 
rules for $L_A$ and consider $\phi$ the parse graph 
representing $\rho_q$, in particular consider the atomic nodes 
representing lexical entries in $\phi$. 
Consider $S$ to be the set of states in $\Phi$ that 
correspond to these nodes. 
Consider simplifying $\Phi$ using techniques that remove 
states and transitions that do not lead to $S$. 
An optimal reduction requires expensive reachability 
analysis, however, we can apply simpler reductions such 
as structural cone of influence reduction,
constant propagation reduction that overapproximate
reachability analysis. 
%For example, we may consider sets of connected states in $\rho_q$ 
%containing a single morpheme as a state in one hierarchy level 
%in $L_A$.
We will perform this simplification before starting the 
morphological analysis since we already have to analyze
the query $q$ and compute $\rho_q$ before we start
the analysis.

{\bf Simplify LCM and MA based on documents.~~}
We will analyze the set of input documents given to a
morphological analyzer component and use the analysis
to further improve the morphological analysis. 
We will extract statistical features from these documents 
to help order the solutions considered by the analyzer
and consider the most probable first.

We will consider the use of HMM~\cite{JanHMM06}
models built on the fly while processing the documents.
A high level HMM model of the document builds an HMM state
model where the HMM states are morphemes or partial 
morphemes and the HMM edges represent transitions between
the HMM states. 
The HMM edges are labeled with weights that reflect the
We will look at building a correspondence between 
the HMM model and the LCM and then simplify the 
LCM where the aggregated probabilities happen to be
below a certain threshold. 
The work in~\cite{JanHMM06} uses a similar approach
to correspond English text to tags corresponding to the 
Unified Medical Language System (UMLS).

We will also consider the use of
latent semantic indexing~\cite{LSI89} techniques. 
We are considering these two techniques as they are 
both language independent and can both be computed and
adapted on the fly.

{\bf Entity and relational extractor.~~}
We will build a set of entity and relational extractors. 
The entities can be proper names, dates, locations,
and qualification adjectives. 
We will explore the use of pairing techniques to establish 
relations between the extracted entities.
The relations will form graphs as in the case studies presented
so far. The nodes represent the entities and the previously 
extracted relations and the edges represent the newly discovered 
relations.

@@

{\bf Statistical components.~~}
We will make use and build statistical components that can be used
in CATARM. 
We will need statistical components to rank solutions, compute 
probable decisions, estimating good thresholds.
